Multivariate Density Estimation: Comparing Transformation Random Forest, Normalizing Flows and Copulas

### 2025-09-29 report/chapter04.Rnw
.

This chapter synthesizes the empirical evidence gathered in Chapter~\ref{ch:dataanalysis}, interprets the behaviour of the estimators within the unified transport frame, and prepares the concluding guidance that follows. We retain the shared preprocessing, likelihood conventions, and diagnostic procedures so that numerical comparisons remain meaningful across synthetic and real datasets.

\section{Interpretation of Results}\label{sec:interpretation-results}
This section interprets the empirical evidence under the unified transport frame. We focus on TRTF-AP, TTM-Sep, and copula baselines evaluated with matched preprocessing, metrics, and units. Synthetic studies report NLL, real datasets report LL, and all values appear in nats with the shared affine correction. These commitments keep objectives, diagnostics, and compute interoperable across estimators.

TRTF-AP often leads within the separable family because additive predictors shift conditional location while the underlying monotone shapes remain stable. The likelihood identities equate TRTF-AP with separable triangular maps, so observed gaps arise from how each estimator realises context shifts and stabilises derivatives. On Half-Moon ($K=2$), TRTF-AP achieved an NLL of $1.71$ while TTM-Sep reached $1.93$, and the first coordinate remained the main source of residual error. Table~\ref{tab:halfmoon-nll} records the per-dimension decomposition and associated uncertainty bands, showing that location adjustments dominate the remaining discrepancies when separability holds approximately in low dimensions.

The four-dimensional generator sharpens this interpretation by isolating coordinates with different conditional structure. TRTF-AP matched the exponential coordinate with an NLL of $1.51$ compared with $1.49$ for the oracle, whereas TTM-Sep over-penalised that coordinate. The beta coordinate produced negative NLLs for the oracles because valid densities can exceed one on $(0,1)$; TRTF-AP approached those values at $-0.25$. The gamma coordinate remained the most challenging, with TRTF-AP at $1.99$ and TTM-Sep at $2.41$. Joint sums favoured TRTF-AP at $4.53$ versus $5.66$, consistent with concentrated gains on location-dominated coordinates. Table~\ref{tab:autoregressive-nll} lists these values, and Figure~\ref{fig:autoregressive-joint-calibration} visualises the residual curvature relative to the identity line.

These comparisons reveal where separability fails to adapt to context-dependent shape. Under a separable map, conditional variance, skewness, and modality remain fixed after the location shift. Probability-integral-transform diagnostics display U-shaped or inverted-U patterns when dispersion misaligns, indicating under- or over-dispersion rather than pure location error. The calibration plots corroborate the per-dimension NLLs and localise remaining structure to the beta and gamma coordinates, where separability is least appropriate. Figure~\ref{fig:autoregressive-joint-calibration} summarises these deviations under the canonical ordering.

Ordering sensitivity stems from finite parameterisations, not from the triangular theory itself. A Knothe--Rosenblatt rearrangement exists for any order, yet limited bases introduce anisotropy that affects fit. Averaging over all $24$ permutations yielded joint NLLs of $4.65$ for TRTF-AP and $5.62$ for TTM-Sep, leaving a $0.97$~nat gap that persisted despite order changes, while the copula baseline averaged $5.45$. Table~\ref{tab:autoregressive-perm} consolidates these permutation-averaged results and underlines the value of data-driven orderings when available.

Small-sample regimes amplified numerical fragility through the log-Jacobian accumulation. TRTF-AP decreased from $38.18$ to $4.64$ joint NLL as $n$ grew from $25$ to $250$, reflecting stabilisation with additional data. TTM-Sep spiked to $6{,}829.45$ at $n=25$ and dropped to $5.61$ at $n=250$, indicating overflow rather than intrinsic misfit. Table~\ref{tab:autoregressive-perm-avg} reports these trajectories, and Section~\ref{sec:models-implementation} documents the derivative clipping and ridge penalties that mitigate this failure mode when samples are scarce.

High dimensionality converts small calibration errors into large likelihood gaps because the triangular determinant accumulates coordinate-wise discrepancies. MINIBOONE with $K=43$ illustrates this accumulation: published flows achieved LL values between $-15.59$ and $-11.68$, whereas TRTF-AP reached $-29.88$ under the shared preprocessing. Table~\ref{tab:uci-loglik} positions TRTF-AP beside the flow baselines and shows that the improvement over the Gaussian reference remains clear even though an approximately $18$~nat gap persists to the strongest flow.

Compute profiles contextualise these accuracy patterns without changing the qualitative ranking at large $K$. At $N=1000$, TRTF-AP required $115$~s on POWER, $138$~s on GAS, $153$~s on HEPMASS, and $202$~s on MINIBOONE, matching the near-linear growth in the training budget and $\mathcal{O}(K)$ evaluation cost. Table~\ref{tab:real-compute} summarises these wall-clock measurements and highlights that separable estimators remain practical in moderate dimensions, yet accuracy dominates the choice once $K \approx 40$.

Taken together, the transport frame delineates when separability suffices and when richer models become necessary. TRTF-AP leads within the separable family when location shifts capture most structure, exhibits ordering sensitivity only through finite bases, and stabilises with modest sample sizes under the safeguards of Section~\ref{sec:models-implementation}. Performance degrades in high dimensions where shape changes and interactions matter, at which point non-separable models offer clear likelihood gains. These conclusions motivate the guidance that will follow in the concluding subsection of this chapter.

\section{Conclusions, Limitations, and Outlook}\label{sec:conclusion-outlook}
We conclude that separable transports remain competitive when conditional location shifts dominate and dimensionality is modest. TRTF-AP led TTM-Sep on Half-Moon ($1.71$ versus $1.93$~NLL) and matched the exponential coordinate in the four-dimensional generator, supporting this interpretation. Conditional decompositions and calibration plots indicate that residual error concentrates in context-dependent shapes, particularly on the beta and gamma components. These findings align with permutation averages that favour TRTF-AP and quantify finite-basis anisotropy. Tables~\ref{tab:halfmoon-nll}--\ref{tab:autoregressive-perm} together with Figure~\ref{fig:autoregressive-joint-calibration} document this evidence under the shared protocol.

Performance on MINIBOONE reveals the cost of separability at higher dimension. TRTF-AP improved the Gaussian reference yet remained about $18$~nats behind the best published flow, consistent with accumulated Jacobian error across $43$ coordinates. POWER exhibited the opposite regime, where TRTF-AP surpassed the reported flow baselines under identical preprocessing. These contrasts suggest that conditional shape and dimension jointly determine whether separable structure suffices. Table~\ref{tab:uci-loglik} reports these comparisons in a common unit.

@@

Several limitations qualify these conclusions. Separable maps fix conditional shape and therefore cannot resolve heteroskedasticity or conditional multimodality. Ordering remained a material source of variance under finite bases, as shown by the $0.97$~nat permutation gap despite stable rankings at moderate sample sizes. Small-sample regimes created numerical fragility through steep log-Jacobian terms, which clipping and ridge regularisation mitigate but do not eliminate. Real-data tables still contain missing GAS and HEPMASS entries, and single-seed settings persist for some runs, limiting external comparability. Tables~\ref{tab:autoregressive-perm}--\ref{tab:real-compute} catalogue these caveats within the standardised protocol.

The outlook follows directly from the evidence. Data-driven orderings are likely to reduce anisotropy without abandoning the lower-triangular map. Low-rank cross-terms in triangular transports and non-additive predictors in TRTF may adapt conditional shapes while preserving monotonicity, exact inversion, and linear per-sample evaluation. Expanded calibration reporting, including probability-integral-transform summaries and Kolmogorov--Smirnov distances, should remain part of any deployment-grade evaluation. Completing GAS and HEPMASS under the same protocol will improve generality and sharpen the accuracy-versus-compute trade-off. These steps target smaller likelihood gaps on high-$K$ datasets while retaining the interpretability and reproducibility provided by the transport frame.

>>>>

This chapter synthesizes the empirical evidence gathered in Chapter~\ref{ch:dataanalysis}, interprets the behavior of the estimators within the unified transport frame, and prepares the concluding guidance that follows. We retain the shared preprocessing, likelihood conventions, and diagnostic procedures so that numerical comparisons remain meaningful across synthetic and real datasets. Copulas enter our study only as low-dimensional ($K\!\le\!3$) diagnostic baselines (e.g., Half-Moon, 4D) and are not evaluated on high-$K$ datasets.

\section{Interpretation of Results}\label{sec:interpretation-results}
This section interprets the empirical evidence under the unified transport frame. We focus on TRTF (additive predictor), TTM-Sep, and, where applicable, copula baselines (only for $K\!\le\!3$) evaluated with matched preprocessing, metrics, and units. Synthetic studies report NLL, real datasets report LL, and we apply the shared affine correction. These commitments keep objectives, diagnostics, and compute interoperable across estimators.

TRTF often leads within the separable family because additive predictors shift conditional location while the underlying monotone shapes remain stable. The likelihood identities equate TRTF with separable triangular maps, so observed gaps arise from how each estimator realizes context shifts and stabilizes derivatives. On Half-Moon ($K=2$), TRTF achieved an NLL of $1.71$ while TTM-Sep reached $1.93$, and the first coordinate remained the main source of residual error. Table~\ref{tab:halfmoon-nll} records the per-dimension decomposition and associated uncertainty bands, showing that location adjustments dominate the remaining discrepancies when separability holds approximately in low dimensions.

The four-dimensional generator sharpens this interpretation by isolating coordinates with different conditional structure. TRTF matched the exponential coordinate with an NLL of $1.51$ compared with $1.49$ for the oracle, whereas TTM-Sep over-penalized that coordinate. The beta coordinate produced negative NLLs for the oracles because valid densities can exceed one on $(0,1)$; TRTF approached those values at $-0.25$. The gamma coordinate remained the most challenging, with TRTF at $1.99$ and TTM-Sep at $2.41$. Joint sums favored TRTF at $4.53$ versus $5.66$, consistent with concentrated gains on location-dominated coordinates. Table~\ref{tab:autoregressive-nll} lists these values, and Figure~\ref{fig:autoregressive-joint-calibration} visualizes the residual curvature relative to the identity line.

These comparisons reveal where separability fails to adapt to context-dependent shape. Under a separable map, conditional variance, skewness, and modality remain fixed after the location shift. Probability-integral-transform diagnostics display U-shaped or inverted-U patterns when dispersion misaligns, indicating under- or over-dispersion rather than pure location error. The calibration plots corroborate the per-dimension NLLs and localize remaining structure to the beta and gamma coordinates, where separability is least appropriate. Figure~\ref{fig:autoregressive-joint-calibration} summarizes these deviations under the canonical ordering.

Ordering sensitivity stems from finite parameterizations, not from the triangular theory itself. A Knothe--Rosenblatt rearrangement exists for any order, yet limited bases introduce anisotropy that affects fit. Averaging over all $24$ permutations yielded joint NLLs of $4.65$ for TRTF and $5.62$ for TTM-Sep, leaving a $0.97$~nat gap that persisted despite order changes, while the copula baseline averaged $5.45$. Table~\ref{tab:autoregressive-perm} consolidates these permutation-averaged results and underlines the value of data-driven orderings when available.

Small-sample regimes amplified numerical fragility through the log-Jacobian accumulation. TRTF decreased from $38.18$ to $4.64$ joint NLL as $n$ grew from $25$ to $250$, reflecting stabilization with additional data. TTM-Sep spiked to $6{,}829.45$ at $n=25$ and dropped to $5.61$ at $n=250$, indicating overflow rather than intrinsic misfit. Table~\ref{tab:autoregressive-perm-avg} reports these trajectories, and Section~\ref{sec:models-implementation} documents the derivative clipping and ridge penalties that mitigate this failure mode when samples are scarce.

High dimensionality converts small calibration errors into large likelihood gaps because the triangular determinant accumulates coordinate-wise discrepancies. MINIBOONE with $K=43$ illustrates this accumulation: published flows achieved LL values between $-15.59$ and $-11.68$, whereas TRTF reached $-30.01$ under the shared preprocessing. Table~\ref{tab:uci-loglik} positions TRTF beside the flow baselines and shows that the improvement over the Gaussian reference remains clear even though an approximately $18$~nat gap persists to the strongest flow.

Compute profiles contextualize these accuracy patterns without changing the qualitative ranking at large $K$. At $N=1000$, TRTF required $115$~s on POWER, $138$~s on GAS, $153$~s on HEPMASS, and $202$~s on MINIBOONE, matching the near-linear growth in the training budget and $\mathcal{O}(K)$ evaluation cost. Table~\ref{tab:real-compute} summarizes these wall-clock measurements and highlights that separable estimators remain practical in moderate dimensions, yet accuracy dominates the choice once $K \approx 40$.

Taken together, the transport frame delineates when separability suffices and when richer models become necessary. TRTF leads within the separable family when location shifts capture most structure, exhibits ordering sensitivity only through finite bases, and stabilizes with modest sample sizes under the safeguards of Section~\ref{sec:models-implementation}. Performance degrades in high dimensions where shape changes and interactions matter, at which point non-separable models offer clear likelihood gains. These conclusions motivate the guidance that will follow in the concluding subsection of this chapter.

\section{Conclusions, Limitations, and Outlook}\label{sec:conclusion-outlook}
We conclude that separable transports remain competitive when conditional location shifts dominate and dimensionality is modest. TRTF led TTM-Sep on Half-Moon ($1.71$ versus $1.93$~NLL) and matched the exponential coordinate in the four-dimensional generator, supporting this interpretation. Conditional decompositions and calibration plots indicate that residual error concentrates in context-dependent shapes, particularly on the beta and gamma components. These findings align with permutation averages that favor TRTF and quantify finite-basis anisotropy. Tables~\ref{tab:halfmoon-nll}--\ref{tab:autoregressive-perm} together with Figure~\ref{fig:autoregressive-joint-calibration} document this evidence under the shared protocol.

Performance on MINIBOONE reveals the cost of separability at higher dimension. TRTF improved the Gaussian reference yet remained about $18$~nats behind the best published flow, consistent with accumulated Jacobian error across $43$ coordinates. POWER exhibited the opposite regime: under identical preprocessing, the reported flows outperformed TRTF (Table~\ref{tab:uci-loglik} lists TRTF at $-7.17$ versus flow baselines near $0.17$ to $0.30$). These contrasts suggest that conditional shape and dimension jointly determine whether separable structure suffices. Table~\ref{tab:uci-loglik} reports these comparisons in a common unit.

@@

Several limitations qualify these conclusions. Separable maps fix conditional shape and therefore cannot resolve heteroskedasticity or conditional multimodality. Ordering remained a material source of variance under finite bases, as shown by the $0.97$~nat permutation gap despite stable rankings at moderate sample sizes. In our $n=250$ synthetic runs, the TRTF versus TTM-Sep ranking did not change across the $24$ permutations (Table~\ref{tab:autoregressive-perm}); ordering affected magnitudes rather than the lead. Simple ordering heuristics (identity or Cholesky-pivoted with optional Gaussianization; see Section~\ref{sec:models-implementation}) reduced variance but did not alter this pattern. Small-sample regimes created numerical fragility through steep log-Jacobian terms, which clipping and ridge regularization mitigate but do not eliminate. Real-data tables still contain missing GAS and HEPMASS entries, and single-seed settings persist for some runs, limiting external comparability. Tables~\ref{tab:autoregressive-perm}--\ref{tab:real-compute} catalog these caveats within the standardized protocol.

The outlook follows directly from the evidence. Data-driven orderings are likely to reduce anisotropy without abandoning the lower-triangular map. Low-rank cross-terms in triangular transports and non-additive predictors in TRTF may adapt conditional shapes while preserving monotone structure, exact inversion, and linear per-sample evaluation. We excluded these richer variants by design in Chapter~\ref{ch:intro} (Non-goals) due to compute and calibration overhead; they remain promising future work once resources permit. Expanded calibration reporting, including probability-integral-transform summaries and Kolmogorov--Smirnov distances, should remain part of any deployment-grade evaluation. Completing GAS and HEPMASS under the same protocol will improve generality and sharpen the accuracy-versus-compute trade-off. These steps target smaller likelihood gaps on high-$K$ datasets while retaining the interpretability and reproducibility provided by the transport frame.

### 2025-09-29 report/chapterA1.Rnw
.

\node[process, font=\small\bfseries, text width=64mm] (data) {Input data: original features $x$};
\node[process, font=\small\bfseries, text width=72mm, below=10mm of data] (std) {Train-only standardization; re-use $\mu$, $\sigma$ from Appendix~\ref{app:ttm-core}};
\node[process, font=\small\bfseries, text width=80mm, below=8mm of std] (jac) {Common pullback evaluation; accumulate $\log\eta(S(u)) + \log|\det \nabla_u S(u)|$ (Equations~\eqref{eq:ch1-pullback}--\eqref{eq:ch1-affine-correction})};

@@

\matrix (ttmMatrix) [below=6mm of ttmTitle, column sep=6mm, row sep=4mm] {
  \node[header, text width=52mm] (ttmMargTitle) {TTM-Marginal}; & \node[header, text width=52mm] (ttmSepTitle) {TTM-Separable}; \\
  \node[process, text width=52mm] (ttmMargStepOne) {Closed-form affine coefficients $(a_k, b_k)$ in $u$-space}; & \node[process, text width=52mm] (ttmSepStepOne) {Context basis $g_k$ on predecessors, monotone basis $h_k$ on $u_k$}; \\
  \node[process, text width=52mm] (ttmMargStepTwo) {Forward pass via \texttt{ttm\_forward}; sum log Jacobian diagonals}; & \node[process, text width=52mm] (ttmSepStepTwo) {Solve constrained optimisation, enforce positivity, store Jacobian terms}; \\
};

@@

@@

  \item \textbf{Prediction (\texttt{predict.mytrtf}):}
    \begin{enumerate}
      \item Convert new data to the same column naming scheme and evaluate $X_1$ through its baseline transformation model to obtain marginal log densities.
      \item For each conditional forest ($k\geq 2$) evaluate the log density of $X_k$ given $X_{1:(k-1)}$, extracting the diagonal when the forest returns a log-density matrix.
      \item Stack the per-dimension log densities (\texttt{logdensity\_by\_dim}) or sum them to obtain the joint log likelihood (\texttt{logdensity}).
    \end{enumerate}
\end{enumerate}

@@

\begin{enumerate}
  \item Provide train-only standardisation helpers that cache feature means and standard deviations and reapply them to new data.
  \item Define basis builders: polynomial features for predecessor coordinates $g_k$, monotone basis functions $f_k$ for the current coordinate, and their derivatives.
  \item Implement optional ordering heuristics (identity or Cholesky pivoting with optional Gaussianisation) and persist selected permutations.
  \item Expose a dispatcher \texttt{ttm\_forward(model, X)} that:
    \begin{enumerate}
      \item Standardises inputs using stored parameters.
      \item For marginal maps applies affine transformations $a_k + b_k x_k$ with precomputed coefficients.
      \item For separable maps constructs $g_k$ and $f_k$, computes $S_k = g_k + f_k$, and records the Jacobian diagonal $\partial_{x_k} S_k$.
    \end{enumerate}
  \item Provide \texttt{ttm\_ld\_by\_dim} to combine the forward map with the Gaussian reference, yielding per-dimension log densities used by all TTM variants.
\end{enumerate}

@@

\begin{enumerate}
  \item Split data into train/test subsets if only a matrix is provided; otherwise accept a prepared list.
  \item Standardise training features and, for each dimension $k$, compute closed-form coefficients $(a_k, b_k)$ that minimise the Gaussian pullback objective subject to $b_k > 0$.
  \item Store model parameters (standardisation, per-dimension coefficients, ordering) and time measurements.
  \item During prediction call \texttt{ttm\_forward} with the marginal coefficients and convert Jacobian diagonals to log densities via \texttt{ttm\_ld\_by\_dim}; aggregate per-dimension contributions when the joint log density is requested.
\end{enumerate}

@@

\begin{enumerate}
  \item Prepare train/test splits and standardise training features as in the marginal case.
  \item For each coordinate $k$:
    \begin{enumerate}
      \item Build polynomial features $g_k$ on previous coordinates (degree set by \texttt{degree\_g}).
      \item Build monotone basis functions $f_k$ on the current coordinate and their derivatives.
      \item If \texttt{degree\_g = 0}, use the marginal closed-form solution to recover affine parameters.
      \item Otherwise solve the regularised optimisation problem
            $\min_c \frac{1}{2}\lVert (I - \Phi_{\text{non}} M)c \rVert^2 - \sum \log (B c) + \lambda\,\text{penalty}(c)$
            using \texttt{optim} with L-BFGS-B while enforcing positivity of the derivative.
      \item Store coefficients $c_{\text{non}}$ and $c_{\text{mon}}$ for the coordinate.
    \end{enumerate}
  \item Assemble the model list with standardisation parameters, coefficients, and metadata; record training/prediction timings.
  \item At prediction time re-use \texttt{ttm\_forward} and \texttt{ttm\_ld\_by\_dim} to obtain per-dimension and joint log densities.
\end{enumerate}

@@

\begin{enumerate}
  \item Define convenience helpers such as \texttt{stderr(x)} and \texttt{add\_sum\_row} for table post-processing.
  \item \texttt{prepare\_data(n, config, seed)} samples from the configured data-generating process, splits the sample into train/validation/test sets, and returns both the matrix of draws and the split structure.
  \item \texttt{fit\_models(S, config)} fits the oracle TRUE density and the TRTF baseline on a split, times their evaluations, and returns the fitted objects together with per-dimension log-likelihood arrays.
  \item \texttt{calc\_loglik\_tables(models, config, X\_te, ...)} aggregates negative log-likelihoods (nats) for TRUE (marginal and joint), TRTF, TTM, and separable TTM, formats the results with standard-error bands, appends a summary row, and renames columns for presentation.
  \item \texttt{eval\_halfmoon(mods, S, out\_csv)} ensures all requisite models are available (TRTF, TTM variants, copula baseline), evaluates them on the half-moon test split, computes joint and per-dimension negative log-likelihoods, and optionally persists the metrics as CSV artefacts.
\end{enumerate}

@@

>>>>

\node[process, font=\small\bfseries, text width=64mm] (data) {Input data: original features $x$};
\node[process, font=\small\bfseries, text width=72mm, below=10mm of data] (std) {Train-only standardization; reuse $\mu$, $\sigma$ from Appendix~\ref{app:ttm-core}};
\node[process, font=\small\bfseries, text width=80mm, below=8mm of std] (jac) {Common pullback evaluation; accumulate $\log\eta(S(u)) + \log|\det \nabla_u S(u)|$ (Equations~\eqref{eq:transport-pullback}--\eqref{eq:transport-affine})};

@@

\matrix (ttmMatrix) [below=6mm of ttmTitle, column sep=6mm, row sep=4mm] {
  \node[header, text width=52mm] (ttmMargTitle) {TTM-Marginal}; & \node[header, text width=52mm] (ttmSepTitle) {TTM-Separable}; \\
  \node[process, text width=52mm] (ttmMargStepOne) {Closed-form affine coefficients $(a_k, b_k)$ in $u$-space}; & \node[process, text width=52mm] (ttmSepStepOne) {Context basis $g_k$ on predecessors, monotone basis $h_k$ on $u_k$}; \\
  \node[process, text width=52mm] (ttmMargStepTwo) {Forward pass via \texttt{ttm\_forward}; sum log Jacobian diagonals}; & \node[process, text width=52mm] (ttmSepStepTwo) {Solve constrained optimization, enforce positivity, store Jacobian terms}; \\
};

@@

% Legend for symbols
\node[header, below=10mm of outputs, text width=120mm] (legendTitle) {Legend};
\node[process, below=2mm of legendTitle, text width=120mm, align=left] (legendBox) {
  $u$: standardized coordinates ($u = T_{\mathrm{std}}(x)$)\\
  $z$: reference coordinates after transport\\
  $S$: lower-triangular transport map ($S:u \mapsto z$)\\
  $\eta$: $K$-variate standard normal density\\
  $\nabla_u S(u)$: Jacobian of $S$ with respect to $u$
};

@@

  \item \textbf{Prediction (\texttt{predict.mytrtf}):}
    \begin{enumerate}
      \item Convert new data to the same column naming scheme and evaluate $X_1$ through its baseline transformation model to obtain marginal log densities.
      \item For each conditional forest ($k\geq 2$) evaluate the log density of $X_k$ given $X_{1:(k-1)}$, extracting the diagonal when the forest returns a log density matrix.
      \item Stack the per-dimension log densities (\texttt{logdensity\_by\_dim}) or sum them to obtain the joint log likelihood (\texttt{logdensity}).
    \end{enumerate}
\end{enumerate}

@@

\begin{enumerate}
  \item Provide train-only standardization helpers that cache feature means and standard deviations and reapply them to new data.
  \item Define basis builders: polynomial features for predecessor coordinates $g_k$, monotone basis functions $f_k$ for the current coordinate, and their derivatives.
  \item Implement optional ordering heuristics (identity or Cholesky pivoting with optional Gaussianization) and persist selected permutations.
  \item Expose a dispatcher \texttt{ttm\_forward(model, X)} that:
    \begin{enumerate}
      \item Standardizes inputs using stored parameters.
      \item For marginal maps apply affine transformations $a_k + b_k x_k$ with precomputed coefficients.
      \item For separable maps constructs $g_k$ and $f_k$, computes $S_k = g_k + f_k$, and records the Jacobian diagonal $\partial_{x_k} S_k$.
    \end{enumerate}
  \item Provide \texttt{ttm\_ld\_by\_dim} to combine the forward map with the Gaussian reference, yielding per-dimension log densities used by all TTM variants.
\end{enumerate}

@@

\begin{enumerate}
  \item Split data into train/test subsets if only a matrix is provided; otherwise accept a prepared list.
  \item Standardize training features and, for each dimension $k$, compute closed-form coefficients $(a_k, b_k)$ that minimize the Gaussian pullback objective subject to $b_k > 0$.
  \item Store model parameters (standardization, per-dimension coefficients, ordering) and time measurements.
  \item During prediction call \texttt{ttm\_forward} with the marginal coefficients and convert Jacobian diagonals to log densities via \texttt{ttm\_ld\_by\_dim}; aggregate per-dimension contributions when the joint log density is requested.
\end{enumerate}

@@

\begin{enumerate}
  \item Prepare train/test splits and standardize training features as in the marginal case.
  \item For each coordinate $k$:
    \begin{enumerate}
      \item Build polynomial features $g_k$ on previous coordinates (degree set by \texttt{degree\_g}).
      \item Build monotone basis functions $f_k$ on the current coordinate and their derivatives.
      \item If \texttt{degree\_g = 0}, use the marginal closed-form solution to recover affine parameters.
      \item Otherwise solve the regularized optimization problem
            $\min_c \frac{1}{2}\lVert (I - \Phi_{\text{non}} M)c \rVert^2 - \sum \log (B c) + \lambda\,\text{penalty}(c)$
            using \texttt{optim} with L-BFGS-B while enforcing positivity of the derivative.
      \item Store coefficients $c_{\text{non}}$ and $c_{\text{mon}}$ for the coordinate.
    \end{enumerate}
  \item Assemble the model list with standardization parameters, coefficients, and metadata; record training/prediction timings.
  \item At prediction time re-use \texttt{ttm\_forward} and \texttt{ttm\_ld\_by\_dim} to obtain per-dimension and joint log densities.
\end{enumerate}

@@

\begin{enumerate}
  \item Define convenience helpers such as \texttt{stderr(x)} and \texttt{add\_sum\_row} for table post-processing.
  \item \texttt{prepare\_data(n, config, seed)} samples from the configured data-generating process, splits the sample into train/validation/test sets, and returns both the matrix of draws and the split structure.
  \item \texttt{fit\_models(S, config)} fits the oracle TRUE density and the TRTF baseline on a split, times their evaluations, and returns the fitted objects together with per-dimension log-likelihood arrays.
  \item \texttt{calc\_loglik\_tables(models, config, X\_te, ...)} aggregates negative log-likelihoods (nats) for TRUE (marginal and joint), TRTF, TTM, and separable TTM, formats the results with standard-error bands, appends a summary row, and renames columns for presentation.
  \item \texttt{eval\_halfmoon(mods, S, out\_csv)} ensures all requisite models are available (TRTF, TTM variants, copula baseline), evaluates them on the half-moon test split, computes joint and per-dimension negative log-likelihoods, and optionally persists the metrics as CSV artifacts.
\end{enumerate}

@@

\subsection{Supplementary Results}\label{app:supplementary}

\begin{table}[htbp]
  \centering
  \textit{(mean NLL in nats).}
  \caption{Permutation-averaged joint test negative log-likelihood for TTM-Sep at $n=25$ on the four-dimensional synthetic generator (aggregated over all $24$ permutations).}
  \label{tab:ttmsep-n25-overflow}
  \begin{tabular}{lr}
    \hline
    Model & $n=25$ \\
    \hline
    TTM-Sep$^{\ast}$ & $6829.45$ \\
    \hline
  \end{tabular}
  \caption*{Note: $^{\ast}$ out-of-scope setting. This value reflects numerical overflow of the separable map in the sparse regime ($n=25$). Stronger derivative clipping and ridge regularization (Section~\ref{sec:models-implementation}) remove this failure in reruns. We mark this configuration as out of scope and exclude it from main-text comparisons; the table remains for transparency.}
\end{table}

\begin{figure}[htbp]
  \centering
  % Simple TikZ markers for ordering sensitivity and mitigation window
  \begin{tikzpicture}[x=28mm,y=18mm]
    % Axes labels
    \node[align=center] at (-0.2,2.2) {Joint NLL\\(lower is better)};
    % TRTF group
    \node[anchor=east] at (-0.1,1.4) {TRTF};
    % TRTF values (from Tables in Chapter 3): min=4.46, canonical=4.53, median=4.59
    \draw[thick,gray!50] (0.0,1.0) -- (0.0,2.0);
    \draw[thick] (0.0,1.23) -- ++(0.12,0) node[anchor=west,scale=0.8] {min $4.46$};
    \draw[thick] (0.0,1.265) -- ++(0.12,0) node[anchor=west,scale=0.8] {canonical $4.53$};
    \draw[thick] (0.0,1.295) -- ++(0.12,0) node[anchor=west,scale=0.8] {median $4.59$};

    % TTM-Sep group
    \node[anchor=east] at (-0.1,0.6) {TTM-Sep};
    % TTM-Sep values: min=5.48, canonical=5.66, median=5.60
    \draw[thick,gray!50] (0.0,0.2) -- (0.0,1.2);
    \draw[thick] (0.0,0.49) -- ++(0.12,0) node[anchor=west,scale=0.8] {min $5.48$};
    \draw[thick] (0.0,0.57) -- ++(0.12,0) node[anchor=west,scale=0.8] {median $5.60$};
    \draw[thick] (0.0,0.63) -- ++(0.12,0) node[anchor=west,scale=0.8] {canonical $5.66$};

    % Brackets indicating potential window for data-driven order selection
    \draw[decorate,decoration={brace,amplitude=5pt}] (-0.02,1.23) -- (-0.02,1.295) node[midway,xshift=-8mm,align=center,scale=0.8] {potential\\gain};
    \draw[decorate,decoration={brace,amplitude=5pt}] (-0.02,0.49) -- (-0.02,0.63) node[midway,xshift=-8mm,align=center,scale=0.8] {potential\\gain};
  \end{tikzpicture}
  \caption{Ordering sensitivity and mitigation window on the four-dimensional generator at $n=250$ (joint NLL; nats; lower is better). Markers show the best (min over $24$ permutations), canonical ordering, and permutation median for each method (values from Chapter~\ref{ch:dataanalysis}: Tables~\ref{tab:autoregressive-nll} and~\ref{tab:autoregressive-perm-spread}). A simple heuristic selects the better of two candidates—identity and Cholesky-pivoted with optional Gaussianization—aiming to move toward the “min” marker while keeping evaluation linear.}
  \label{fig:ordering-heuristics-4d}
\end{figure}

% Clipping bounds are reported inline per study in Chapter~\ref{ch:dataanalysis} and stored with the experiment logs; we omit a dedicated summary table to avoid duplication.
